Protected Areas Practical

### 2024-11-22 tutorial/MSc_PA_practical_excercise.Rmd
added readme info

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)#
#  rmarkdown::html_document:
#    theme: lumen
  # prettydoc::html_pretty:
  #   theme: hpstr
  #   highlight: github
```

@@

1. Go to https://github.com/KiranLDA/MSc_Protected_Areas
2. Click on the big green button 'CODE'.
3. Select 'Download ZIP'.
4. Save the file to your desktop.
5. Once opened unzip/extract all files in the ZIP file.

@@

# Excercise 1

@@

@@

```{r species}
# put them together into a multi-layer raster
species = terra::rast(paste0(basepath,"species.tif"))

@@

A very important element of protected area planning is the cost. Wherever we put a protected, there will be a cost involved. Cost can be direct (someone has done an economic evaluation about economic losses of converting land from e.g. grazing land to  protected land, and the costs of purchasing that land). However, costs are more often indirectly estimated using, for example, the human footprint index, to see where people are using land. We then make the assumption that areas that people use are most costly and created the greatest conflict, and should be avoided as a place to put a protected area.

```{r cost}
# load the file
cost0 = rast(paste0(basepath, "cost.tif"))

@@

```{r image_grobs, fig.show = "hold", out.width = "50%", fig.align = "default", echo=FALSE}

@@

## How many were not protected 30%
```{r}
print(mean(p1_target_coverage$met) * 100)
```
# Excercise 1

@@

```{r}
# load protected areas
PAs = terra::rast(paste0(basepath, "locked_in.tif"))

@@

```{r}
# need to reset budget because it was modified in exercise 1
budget <- terra::global(cost, "sum", na.rm = TRUE)[[1]] * 0.3

@@

# Exercise 2

@@

# Exercise 3 (Bonus)

@@

```{r, fig.width=5, fig.height=7, echo=FALSE}
locked_out = terra::rast(paste0(basepath, "locked_out.tif"))

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(root.dir = "C:/Users/kdh10kg/Documents/github/MSc_Protected_Areas/",
                      echo = TRUE)#
```

@@

1. Go to https://github.com/KiranLDA/MSc_Protected_Areas
2. Click on the big green button 'CODE'.
3. Select 'Download ZIP'.
4. Save the file to your desktop.
5. Once opened unzip/extract all files in the ZIP file.
6. Then also unzip the rasters inside `data/data.zip` into the `data` folder.

@@

# Exercise 1

@@

# Start the practical in R

Now get the code for the practical from Go to `code/02_practical.R`

# Set the Working Directory

Replace with the path where you downloaded `MSc_Protected_Areas`. This means that your code will run from this directory. Make sure you use `/` or `\\` in your path name, otherwise you will get an error.

```{r wd}
setwd("C:/Users/kdh10kg/Documents/github/MSc_Protected_Areas/")
```

@@

```{r species}
# put them together into a multi-layer raster
species = terra::rast("data/species.tif")
```

@@

A very important element of protected area planning is the cost. Wherever we put a protected area, there will be a cost involved. Cost can be:

- directly estimated: we can do an economic evaluation about economic losses of converting land to  protected land. This can be through the cost of purchasing land, costing the staff needed to patrol, the revenue from tourism, the loss of opportunities (loss of mining land, grazing land, agricultural land, and even homes).
- indirectly estimated: we can estimate what land is most costly. For example, the human footprint index is often used to see where people are using land. We then make the assumption that areas that people use are most costly and created the greatest conflict, and should be avoided as a place to put a protected area.

Here, it's an excecise, so we use the Human footprint layer, which is downloaded from this paper:
https://www.nature.com/articles/s41597-022-01284-8#Sec12

```{r cost}
# load the file
cost0 = rast("data/cost.tif")

@@

```{r image_grob8, fig.show = "hold", out.width = "40%", fig.align = "default", echo=FALSE}

@@

## How many were not protected 30%
```{r}
print(mean(p1_target_coverage$met) * 100)
```
# Exercise 2

@@

```{r}
# load protected areas
PAs = terra::rast("data/locked_in.tif")

@@

```{r}
# need to reset budget because it was modified in exercise 2
budget <- terra::global(cost, "sum", na.rm = TRUE)[[1]] * 0.3

@@

# Exercise 3

@@

# Exercise 4 (Bonus)

@@

```{r, fig.width=5, fig.height=7, echo=FALSE}
locked_out = terra::rast("data/locked_out.tif")
